### 2020-09-12 paper/experiments.tex
Minor fixes

\subsection{Experiments}
\label{sec:experiments}

@@

When looking at the results (\Cref{plot:perfLines}), while na\"ive algorithm's time is almost zero, we ser the CCLSV running time climbing with the growth of $|V|$, with only moderate improvement by GPU CCLSV. Slight GPU improvement is explained by anylysis of the components of the overall time (\Cref{plot:perfDet}). Testing all near cleaners takes up around half of total time, so we cannot speed up the overall time by much. These are also relatively small tests, so the latency from copying data to the GPU is significant.

@@

Next, we turn our attention to graphs generated on a grid, or a checker board (see later paragraphs). First, we take lattice graphs. In lattice graph, each vertex is connected to vertices that are above, below, to the left and to the right of it, if such vertices exist. See \cref{fig:lattice} for an example of $4 \times 4$ lattice.

@@

With growing $|V|$ also grows the speedup gained from utilizing GPU. For $|V| = 78$ the running time for the whole algorithm is almost 3x better.

@@

Rook graphs have many more chords than Lattice graphs, so the na\"ive's performance is much better here. What is more interesting, the overall performance of CCLSV is much worse (\cref{plot:rookLines}). Looking at \cref{fig:rookDet} we can see that the majority of CCLSV time is spent checking all near cleaners. This can be explained by the fact that rook graphs are much denser than lattice graphs and have many more 3-vertex paths (see \cref{alg:testNearCleaner}). This is confirmed by greater CCLSV GPU speedup of around 6x on graph with $|V| = 35$.

@@

Hypercube graphs tell a similar story to lattice graphs and knight graphs. Na\"ive is very fast at first, but becomes unbearably slow for $|V|$ higher than 55. Unfortunately, CCLSV doesnt fare much better, although it grows slower. Again, the speedup of GPU CCLSV is noticeable, being around 4.8x for $|V| = 55$.

@@

There are no long chordless paths here, so na\"ive fares great (\cref{plot:splitLines}). What is interesting, when looking at detailed times of GPU CCLSV, we can see that checking the existence of simple forbidden structures takes more than 30\% of the time. This opens a possibility of further GPU speedups, as previous tests indicated that checking simple forbidden structures is usually very quick on the CPU.

@@

Calculating $\vartheta(G)$ is the most complicated part of the coloring algorithm. With that done by an external library, the rest of the program is a straightforward implementation of the algorithms in \cref{sec:coloringEllipsoid}. Our main goal of the implementation was to check if this method is still impractical, even on modern equipment. Let us proceed straight to experiments and results.

@@

A few things of note. We notice that in rook graphs the graph manipulation other than calculating CSDP takes close to half of overall running time (\cref{plot:colorRookDet}). We didn't particularly optimize this, other than by use of methods developed for Berge graph recognition algorithm. There are surely some optimizations to be made here if need be. Altough, as we can see on other plots, it won't help im majority of the cases.

\subsection{Combinatorial Method}

Cite the paper.\\

On its complexity - point to appendix for pseudo-code.

>>>>

\section{Experiments}
\label{sec:experiments}

@@

When looking at the results (\Cref{plot:perfLines}), while na\"ive algorithm's time is almost zero, we see the CCLSV running time climbing with the growth of $|V|$, with only moderate improvement by GPU CCLSV. Slight GPU improvement is explained by anylysis of the components of the overall time (\Cref{plot:perfDet}). Testing all near cleaners takes up around half of total time, so we cannot speed up the overall time by much. These are also relatively small tests, so the latency from copying data to the GPU is significant.

@@

Next, we turn our attention to graphs generated on a grid, or a checker board (see later paragraphs). First, we take lattice graphs. In lattice graph, each vertex is connected to vertices that are above, below, to the left and to the right of it, if such vertices exist. See \cref{fig:lattice} for an example of a $4 \times 4$ lattice graph.

@@

With larger $|V|$ the speedup gained from utilizing GPU is also more significant. For $|V| = 78$ the running time for the whole algorithm is almost 3x better.

@@

Rook graphs have many more chords than Lattice graphs, so the na\"ive's performance is much better here. What is more interesting, the overall performance of CCLSV is much worse (\cref{plot:rookLines}). Looking at \cref{plot:rookDet} we can see that the majority of CCLSV time is spent checking all near cleaners. This can be explained by the fact that rook graphs are much denser than lattice graphs and have many more 3-vertex paths (see line TODO of \cref{alg:testNearCleaner}). This is confirmed by greater CCLSV GPU speedup of around 6x on graphs with $|V| = 35$.

@@

Hypercube graphs tell a similar story to lattice graphs and knight graphs. Na\"ive is very fast at first, but becomes unbearably slow for $|V|$ higher than 55. Unfortunately, CCLSV doesn't fare much better, although it grows slower. Again, the speedup of GPU CCLSV is noticeable, being around 4.8x for $|V| = 55$.

@@

There are no long chordless paths here, so na\"ive fares great (\cref{plot:splitLines}). What is interesting, when looking at detailed times of GPU CCLSV, we can see that checking the existence of simple forbidden structures takes more than 30\% of the time. This opens a possibility of further GPU speedups, not explored by us.

@@

Calculating $\vartheta(G)$ is the most complicated part of the coloring algorithm. With that done by an external library, the rest of the program is a straightforward implementation of the algorithms in \cref{sec:coloringEllipsoid}. In most of our tests, the majority of running time was consumed on calculating $\vartheta$ of various graphs. As it was done by an external library, there isn't much optimization potential for us. One thing of note is, that specifying the precision of $\vartheta$ we want to calculate to be $1/3$ sped up the algorithm by TODO. Our main goal of the implementation was to check if this method is still impractical, even on modern equipment. Let us proceed straight to experiments and results.

@@

A few things of note. We notice that in rook graphs the graph manipulation other than calculating CSDP takes close to half of overall running time (\cref{plot:colorRookDet}). We didn't particularly optimize this, other than by use of methods developed for Berge graph recognition algorithm. There are surely some optimizations to be made here if need be. Altough, as we can see on other plots, it won't help in majority of the cases.

### 2020-09-12 paper/implementation.tex
Minor fixes

The Berge recognition algorithm's running time is $O(n^9)$, which brings into question its applicability to any real use case. Although time complexity is indeed a limiting factor, a number of lower level optimizations done on implementation's level (\cref{sec:Optimizations}) make a very big difference and make it usable -- or at least much more usable than na\"ive algorithm could be. Also, we explore a new frontier of implementing its most time consuming part on massively parallel GPU architecture, with good results (\cref{sec:CUDA}).

@@

When implementing a complicated algorithm that has a time complexity of $O(n^9)$ optimizations can be both crucial and difficult to implement. There is not a single code path that takes up all the running time -- or at lest there isn't one from theoretical point of view. Therefore a tool for inspecting running time bottlenecks is needed. We used Valgrind's tool called \emph{callgrind} \cite{callgrind}.

@@

\TODO{bitsets}

@@

\subsection{Parallelism with CUDA (?)}
\label{sec:CUDA}

\TODO{a section?}
\TODO{CUDA background}
\TODO{moderngpu - allows us to run simple transforms}
\TODO{Implementing get NC is tough -- we rely on set<bitset> to do the work. Maybe we could use a dynamic set from Adam's link? As a "open question"}

We posed a question if it would be profitable to use GPU for this problem. On one hand the graphs we are working on are small and our time complexity so large that a speedup from massively parallel architecture could be profitable. On the other the algorithm is complex and not easy to parallelise. Therefore we decided to implement whole na\"ive algorithm and most time consuming parts of the CCLSV\todo{a good name for it?} algorithm.

\subsubsection{Na\"ive parallelisation}

Na\"ive algorithm is very simple -- generate all possible odd hole candidates, check if any of them is an odd hole and repeat for the complement. But its parallelisation is not trivial.

The biggest problem of implementing na\"ive algorithm is in splitting the work between CUDA threads. This is very important because while GPU is faster when all its cores are working efficiently, single core performance is much slower than CPU. But it is not obvious how to split the work.

Our first attempt was to switch our slightly more sophisticated way of generating paths (\cref{sec:usesGeneration}) in favor of generating all combinations. We could code a prefix of a combination as a number (we use combinations with repetitions for simplicity). This would give us $|V(G)|^k$ codes for all prefixes of length $k$. Then each CUDA thread would process its part of all codes: for each code it would first check if the encoded prefix is a valid path and then generate all path candidates with that prefix. This approach has two problems: it doesn't split the work evenly enough and it does too much unneeded work (only a few prefix codes are valid). Still, implementing it on GPU gives us a program with speeds comparable to na\"ive CPU algorithm after all optimizations.

A better approach is to generate all paths of some length $k$ on CPU (using algorithm described in \cref{sec:usesGeneration}), copy them on GPU and have each thread process them as described above. This is superior to the previous algorithm in utilizing GPU -- each prefix takes similar time to process. We also even out the work done by each thread by having single thread process multiple prefixes. But it puts much more strain on the CPU -- it has to generate all valid prefixes. Here we used experiments to determine best value of $k$ (it was 7 for graphs of $n < 15$ \todo{check} and 6 for larger). This yields an algorithm that beats CPU na\"ive by a factor of 50x. \todo{check}.

\subsubsection{CCLSV parallelisation}

@@

Let us recall the \cref{alg:testNearCleaner}. It calculates array $R$ of shortest paths, then for each 3-vertex path and an additional vertex it does $O(1)$ checks to see if they along with two paths from $R$ give us an odd hole. We will parallelise the work after calculating $R$. Let us notice, that for all $X$s all 3-vertex paths are the same. We calculate them beforehand \todo{do it also in CPU} \todo{acutally do this}. Then each thread receives a 3-vertex path and an additional vertex and performs the required checks. This is almost perfect scenario for GPU -- we do a simple SIMD work in parallel, without a lot scattered memory access.

It turns out that this optimization alone speeds up \cref{alg:testNearCleaner} almost 30x \todo{check} on bigger tests, which gives us a speedup of about 5x \todo{check} for overall algorithm.

>>>>

The Berge recognition algorithm's running time is $O(n^9)$, which brings into question its applicability to any real use case. Although time complexity is indeed a limiting factor, a number of lower level optimizations done on implementation's level (\cref{sec:Optimizations}) make a very big difference and make it more viable than na\"ive algorithm, at least on some test cases. Also, we explore a new frontier of implementing its most time consuming part on massively parallel GPU architecture, with some good results (\cref{sec:CUDA}).

@@

When implementing a complicated algorithm that has a time complexity of $O(n^9)$ optimizations can be both crucial and difficult to implement. There is not a single code path that takes up all the running time -- or at lest there isn't one from a theoretical point of view. Therefore a tool for inspecting running time bottlenecks is needed. We used Valgrind's tool called \emph{callgrind} \cite{callgrind}.

@@

Another algorithm for which we found major speedups is the algorithm to generate all possible near-cleaners (\Cref{alg:listNearCleaners}). At its very end, it returns a set that is an union \todo{after rewriting alg, point to a line}. We used a \texttt{dynamic\_bitset} from the boost library \cite{boost} for this purpose. It is a data structure to represent a set of bits, that also allows for fast bitwise operators that one can apply to builtin integers. By using it, the speedup of \Cref{alg:listNearCleaners} was about x. \todo{todo}

After these optimizations, checking each potential near-cleaner (\Cref{alg:testNearCleaner}) is by far the biggest bottlenecks of the CCLSV algorithm. We didn't find any major optimizations there, but \Cref{alg:testNearCleaner} has a good potential for parrallelization, which we explore in \Cref{sec:CUDA}.

@@

\section{Parallelism with CUDA}
\label{sec:CUDA}

\TODO{CUDA background}
\TODO{moderngpu - allows us to run simple transforms}
\TODO{Implementing get NC is tough -- we rely on set<bitset> to do the work. Maybe we could use a dynamic set from Adam's link? As a "open question"}

We posed a question if it would be profitable to use GPU for this problem. On one hand the graphs we are working on are small and our time complexity so large that a speedup from massively parallel architecture could be profitable. On the other the algorithm is complex and not easy to parallelise. Therefore we decided to analyze the  CCLSV\todo{a good name for it?} algorithm and parrallelize parts that would benefit from it the most.

% \subsubsection{Na\"ive parallelisation}

% Na\"ive algorithm is very simple -- generate all possible odd hole candidates, check if any of them is an odd hole and repeat for the complement. But its parallelisation is not trivial.

% The biggest problem of implementing na\"ive algorithm is in splitting the work between CUDA threads. This is very important because while GPU is faster when all its cores are working efficiently, single core performance is much slower than CPU. But it is not obvious how to split the work.

% Our first attempt was to switch our slightly more sophisticated way of generating paths (\cref{sec:usesGeneration}) in favor of generating all combinations. We could code a prefix of a combination as a number (we use combinations with repetitions for simplicity). This would give us $|V(G)|^k$ codes for all prefixes of length $k$. Then each CUDA thread would process its part of all codes: for each code it would first check if the encoded prefix is a valid path and then generate all path candidates with that prefix. This approach has two problems: it doesn't split the work evenly enough and it does too much unneeded work (only a few prefix codes are valid). Still, implementing it on GPU gives us a program with speeds comparable to na\"ive CPU algorithm after all optimizations.

% A better approach is to generate all paths of some length $k$ on CPU (using algorithm described in \cref{sec:usesGeneration}), copy them on GPU and have each thread process them as described above. This is superior to the previous algorithm in utilizing GPU -- each prefix takes similar time to process. We also even out the work done by each thread by having single thread process multiple prefixes. But it puts much more strain on the CPU -- it has to generate all valid prefixes. Here we used experiments to determine best value of $k$ (it was 7 for graphs of $n < 15$ \todo{check} and 6 for larger). This yields an algorithm that beats CPU na\"ive by a factor of 50x. \todo{check}.

% \subsubsection{CCLSV parallelisation}

@@

Let us recall the \cref{alg:testNearCleaner}. It calculates array $R$ of shortest paths, then for each 3-vertex path and an additional vertex it does $O(1)$ checks to see if they along with two paths from $R$ give us an odd hole. We will parallelise the work after calculating $R$. Let us notice, that for all $X$s all 3-vertex paths are the same. We calculate them beforehand, then each thread receives a 3-vertex path $x_1 - x_2 - x_3$ and an additional vertex $y_1$ and performs the required checks. This is almost perfect scenario for the GPU -- we do a simple SIMD work in parallel, without a lot scattered memory access.

It turns out that this optimization alone speeds up \cref{alg:testNearCleaner} almost 7x \todo{check} on bigger tests, which gives us a speedup of about 6x \todo{check} for overall algorithm.
